Parallel Tractability of Ontology\\ Materialization: Technique and Practice

### 2018-01-31 parallel-tractability-J-v9/background.tex
Section 4 and small adjustments done

We discuss the main issues in this paper using standard datalog notions.
In datalog \cite{database}, a \emph{term} is a variable or a constant. An \emph{atom} $A$
is defined by $A\equiv p(t_1,...,t_n)$ where $p$ is a \emph{predicate} (or \emph{relational})
name, $t_1,...,t_n$ are terms, and $n$ is the arity of $p$. If all the terms in an atom $A$ are
constants, then $A$ is called a \emph{ground atom}.
A datalog \emph{rule} is of the form: `$B_1,...,B_n\rightarrow H$',\footnote{In datalog rules, a comma
represents a Boolean conjunction `$\wedge$'.} where $H$ is referred to as
the \emph{head atom} and $B_1,...,B_n$ the \emph{body atoms}. Each variable in the head atom
of a rule must occur in at least one body atom of the same rule. A \emph{fact} is a rule of
the form `$\rightarrow H$', i.e., a rule with an empty body and the head $H$ being a ground atom.
A datalog program $P$ consists of rules and facts.
A \emph{substitution} $\theta$ is a partial mapping of variables to constants.
For an atom $A$, $A\theta$ is the result of replacing each variable $x$ in $A$
with $\theta(x)$ if the latter is defined. We call $\theta$ a \emph{ground substitution}
if each defined $A\theta$ is a ground atom.
A \emph{ground instantiation} of a rule is obtained by applying a ground substitution on all
the terms in this rule with respect to a finite set of constants occurring in $P$.
Furthermore the ground instantiation of $P$, denoted by $P^*$,
consists of all ground instantiations of rules in $P$.
The predicates occurring only in the body  of some rules are called \emph{EDB predicates},
while the predicates that may occur as head atoms are called \emph{IDB predicates}.

@@

In the initial work of DHL \cite{GrosofHVD03}, \emph{complex role inclusion axioms} (complex RIAs) of
the form $R_1\circ...\circ R_n\sqsubseteq R$ are not considered, although they can be
naturally transformed into datalog rules.
In this paper, we also consider an extension of DHL (denoted by DHL($\circ$))
that allows complex RIAs. Since a complex RIA can be transformed to
several axioms of form (R4), we then require that an
RBox $\mathcal{R}$ of a DHL($\circ$) ontology can contain
axioms of the forms (R1-R4). Note that (R3) is actually a special
case of (R4).

@@

An ontology
expressed in DHL, DHL($\circ$), DL-Lite$_{\text{core}}$ or DL-Lite$_{\mathcal{R}}$ can be transformed into a datalog program
(see the corresponding rules in the right column of Table~1).
Note that an axiom of form (T4) $A\sqsubseteq\exists R$ should be
transformed into a
first-order logic rule of the form $A(x)\rightarrow \exists y(R(x,y))$, where $y$ is called a \emph{free variable} (it is
not occurring in the body atom $A(x)$).
In this work, such a free variable $y$ is eliminated via Skolemisation into a fresh individual $o_{R}^A$
that corresponds to the concept $A$ and the role $R$. The elimination of free variables
does not influence the result of materialization \cite{CalvaneseGLLR07}.

@@

Based on the above representations, ontology materialization
corresponds to the evaluation of datalog programs.
Specifically, given a datalog program $P=\langle R, \textbf{I}\rangle$,
let $T_R(\textbf{I})=\{H\theta|\forall B_1,...,B_n\rightarrow H\in R,
B_i\theta\in\textbf{I}, 1\leq i\leq n \}$,
where $\theta$ is some substitution; further let $T_R^{0}(\textbf{I})=\textbf{I}$ and
 $T_R^{i}(\textbf{I})=T_R^{i-1}(\textbf{I})\cup T_R(T_R^{i-1}(\textbf{I}))$ for each $i>0$.
The smallest integer $n$ such that $T_R^{n}(\textbf{I})= T_R^{n+1}(\textbf{I})$ is called \emph{stage},
and \emph{materialization} refers to the computation of $T_R^{n}(\textbf{I})$ with respect to $R$ and \textbf{I}.
$T_R^{n}(\textbf{I})$ is also called the \emph{fixpoint} and denoted by $T_R^{\omega}(\textbf{I})$.
We say that an atom $A$ is \emph{derivable} or can be \emph{derived} with respect
to the datalog program $P=\langle R, \textbf{I}\rangle$ if $A\in T_R^{\omega}(\textbf{I})$.
In this paper, we consider the data complexity of materialization, i.e., we \emph{assume that the rule set $R$ is fixed}
for any class of datalog programs.

>>>>

We discuss the main issues in this paper using standard datalog notions.
In datalog \cite{database}, a \emph{term} is a variable or a constant. An \emph{atom} $A$
is defined by $A\equiv p(t_1,\ldots,t_n)$ where $p$ is a \emph{predicate} (or \emph{relational})
name, $t_1,\ldots,t_n$ are terms, and $n$ is the arity of $p$. If all the terms in an atom $A$ are
constants, then $A$ is called a \emph{ground atom}.
A datalog \emph{rule} is of the form: `$B_1,\ldots,B_n\rightarrow H$',\footnote{In datalog rules, a comma
represents a Boolean conjunction `$\wedge$'.} where $H$ is referred to as
the \emph{head atom} and $B_1,\ldots,B_n$ the \emph{body atoms}. Each variable in the head atom
of a rule must occur in at least one body atom of the same rule. A \emph{fact} is a rule of
the form `$\rightarrow H$', i.e., a rule with an empty body and the head $H$ being a ground atom.
A datalog program $P$ consists of rules and facts.
A \emph{substitution} $\theta$ is a partial mapping of variables to constants.
For an atom $A$, $A\theta$ is the result of replacing each variable $x$ in $A$
with $\theta(x)$ if the latter is defined. We call $\theta$ a \emph{ground substitution}
if each defined $A\theta$ is a ground atom.
A \emph{ground instantiation} of a rule is obtained by applying a ground substitution on all
the terms in this rule with respect to a finite set of constants occurring in $P$.
Furthermore the ground instantiation of $P$, denoted by $P^*$,
consists of all ground instantiations of rules in $P$.
The predicates occurring only in the body  of some rules are called \emph{EDB predicates},
while the predicates that may occur as head atoms are called \emph{IDB predicates}.

@@

In the initial work of DHL \cite{GrosofHVD03}, \emph{complex role inclusion axioms} (complex RIAs) of
the form $R_1\circ\ldots\circ R_n\sqsubseteq R$ are not considered, although they can be
naturally transformed into datalog rules.
In this paper, we also consider an extension of DHL (denoted by DHL($\circ$))
that allows complex RIAs. Since a complex RIA can be transformed to
several axioms of form (R4), we then require that an
RBox $\mathcal{R}$ of a DHL($\circ$) ontology can contain
axioms of the forms (R1-R4). Note that (R3) is actually a special
case of (R4).

@@

An ontology
expressed in DHL, DHL($\circ$), DL-Lite$_{\text{core}}$ or DL-Lite$_{\mathcal{R}}$ can be transformed into a datalog program
(see the corresponding rules in the right column of Table~1).
Note that an axiom of form (T4) $A\sqsubseteq\exists R$ should be
transformed into a
first-order logic rule of the form $A(x)\rightarrow \exists y(R(x,y))$, where $y$ is called a \emph{free variable} (it is
not occurring in the body atom $A(x)$).
In this work, such a free variable $y$ is eliminated via Skolemization into a fresh individual $o_{R}^A$
that corresponds to the concept $A$ and the role $R$. The elimination of free variables
does not influence the result of materialization \cite{CalvaneseGLLR07}.

@@

Based on the above representations, ontology materialization
corresponds to the evaluation of datalog programs.
Specifically, given a datalog program $P=\langle R, \textbf{I}\rangle$,
let $T_R(\textbf{I})=\{H\theta|\forall B_1,\ldots,B_n\rightarrow H\in R,
B_i\theta\in\textbf{I}, 1\leq i\leq n \}$,
where $\theta$ is some substitution; further let $T_R^{0}(\textbf{I})=\textbf{I}$ and
 $T_R^{i}(\textbf{I})=T_R^{i-1}(\textbf{I})\cup T_R(T_R^{i-1}(\textbf{I}))$ for each $i>0$.
The smallest integer $n$ such that $T_R^{n}(\textbf{I})= T_R^{n+1}(\textbf{I})$ is called \emph{stage},
and \emph{materialization} refers to the computation of $T_R^{n}(\textbf{I})$ with respect to $R$ and \textbf{I}.
$T_R^{n}(\textbf{I})$ is also called the \emph{fixpoint} and denoted by $T_R^{\omega}(\textbf{I})$.
We say that an atom $A$ is \emph{derivable} or can be \emph{derived} with respect
to the datalog program $P=\langle R, \textbf{I}\rangle$ if $A\in T_R^{\omega}(\textbf{I})$.
In this paper, we consider the data complexity of materialization, i.e., we \emph{assume that the rule set $R$ is fixed}
for any class of datalog programs.
